---
title: "Geoms and Aesthetics"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Plotting_In_R/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---

<p>&nbsp;</p>
These exercises cover the Geoms and Aesthetics of ggplot2 for [Plotting in R](https://rockefelleruniversity.github.io/Plotting_In_R/). 

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```


These first few exercises will run through some of the simple principles of creating a ggplot2 object, assigning aesthetics mappings and geoms.

**Exercise 1** - **Scatterplots**

- Read in the cleaned patients dataset as we saw in ggplot2 course earlier ("patients_clean_ggplot2.txt")

```{r,echo=toMessage}
getwd()
patients_clean <- read.delim("./data/patients_clean_ggplot2.txt",sep="\t")
```


- Using the patient data set generate a scatter plot of BMI versus Weight.

```{r,echo=toMessage}
library(ggplot2)

plot <- ggplot(data=patients_clean,
               mapping=aes(x=BMI,y=Weight))+geom_point()
plot
```

- Extend the plot by adding a color scale to the scatter plot based on the Height variable.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(x=BMI,y=Weight,colour=Height))+geom_point()
plot
```


- Using an additional geom, add an extra layer of a fit line to our scatterplot.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(x=BMI,y=Weight,colour=Height))+geom_point()+
  geom_smooth()
plot
```

- Does the fit in look good? Look at the description for ?geom_smooth() and adjust the method for a better fit.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(x=BMI,y=Weight,colour=Height))+geom_point()+
  geom_smooth(method="lm",se=F)
plot
```

**Exercise 2** - **Boxplots and Violin plots**

- Generate a boxplot of BMIs comparing smokers and non-smokers.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(x=Smokes,y=BMI))+geom_boxplot()
plot
```


- Following from the boxplot comparing smokers and non-smokers, color boxplot edges by Sex.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(x=Smokes,y=BMI,color=Sex))+geom_boxplot()

plot
```

**Exercise 3** - **Histogram and Density plots**

- Generate a histogram of BMIs with each bar colored blue.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(BMI))+
  geom_histogram(fill="blue")
plot
```

- Generate density plots of BMIs colored by Sex. Make sure you can see both distributions by adjusting the transparency. 

HINT: alpha can be used to control transparency.

```{r,echo=toMessage}

plot <- ggplot(data=patients_clean,
               mapping=aes(BMI))+ geom_density(aes(fill=Sex), alpha=0.5)
plot
```




