---
title: "Themes, Saving and Interactive Plots"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/Plotting_In_R/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false  
---

<p>&nbsp;</p>
These exercise cover the scales, statistics and themes of ggplot2 for [Plotting in R](https://rockefelleruniversity.github.io/Plotting_In_R/). 


```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

**Exercise 1** - **Themes**

- Read in the cleaned patients dataset as we saw in ggplot2 course earlier ("patients_clean_ggplot2.txt")

```{r,echo=toMessage}
patients_clean <- read.delim("data/patients_clean_ggplot2.txt",sep="\t")
```

-  Using the patient data set generate a scatter plot of BMI versus Weight. Add a color scale to the scatter plot based on the Height variable. Use an additional geom, add an extra layer of a fit line to our scatterplot (use lm method).


```{r,echo=toMessage}

library(ggplot2)
plot <- ggplot(data=patients_clean,
               mapping=aes(x=BMI,y=Weight,colour=factor(Age)))+geom_point()+stat_smooth(method="lm",se=F)
plot
```

- Remove the legend title from the plot, change the background colors of legend panels to white and place legend at bottom of plot.

```{r,echo=toMessage}

plot <- plot+theme(legend.title=element_blank(),legend.background=element_rect(fill="white"),legend.key=element_rect(fill="white"),legend.position="bottom")

plot
```

- Add a title to the plot, remove minor grid lines and save to 7 by 7 inch png file.

```{r,echo=toMessage}

plot <- plot+ggtitle("BMI vs Weight")+theme(panel.grid.minor=element_blank())

plot

ggsave(plot,file="BMIvsWeight.png",units = "in",height = 7,width = 7)
```

- Produce a Height vs Weight scatter plot with point sizes scaled by BMI. 
Present only the points and title of plot with all other graph features missing.

